Lab 6 - Diagnostics and simulation based inferences: Answers

### 2023-01-04 Lab 3  Fitting LMs/Lab3_LM_Answers.Rmd
Add files via upload

---
title: "Lab 3: The Linear Model in R"
author: "Eric Parajon"
output: pdf_document
---
Adapted  from code and text from Isabel Laterzo, Simon Hoellerbauer and Chelsea Estancona.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, eval=FALSE)

@@

```{r, eval = FALSE}
set.seed(123)

##Explanatory variables
x1 <- rnorm(__, __, __)

@@

error <- rnorm(length(x1), 0, sigma) #Let's simulate some random error, too.

What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r, eval = FALSE}
y <- _____________________
```

@@

```{r, eval = FALSE}
plot(x1, y)
```

Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. 
The basic syntax for this function is `lm(Y ~ X1 + X2)`.

Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

```{r, eval = FALSE}
m1 <- lm(___________) #the basic syntax here: lm(Y ~ X1 + X2)
#summarize the 'model object' we've created
summary(m1)
```

@@

We can also combine what we learned about using the tidyverse with linear modeling!

As we dive into learning linear models in `R`, we're going to use some data from the `mtcars`. But first, lets use the `tidyverse` to clean it up a bit. First, take a look at the mtcars data (its readily available so you don't really need to load it in!).

```{r}
library(tidyverse)
#examine data
summary(mtcars)
```

@@

Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).
```{r, eval = FALSE}

new_data <- mtcars %>%
  filter(cyl > 4) %>%
  select(__, __, __)
```

Now that we have our data ready, we can use `lm()`.  Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m2` in this case) in order to see some of the most interesting information about the fitted model.

@@

```{r, eval = FALSE}
#model
m2 <- lm(___________)

#summarize
summary(___)

@@

```{r, eval=FALSE, echo=TRUE}
#Taking a look at m1
____________

#extract coefficients
m2$_____

#extract residuals
m2$______

#extract y hat 
m2$_______
```

@@

```{r, eval = FALSE}
m2$coefficients[1] #What does this give us?
m2[2] #What about this?
```

What if we wanted to include an interaction term? Try interacting `disp` and `cyl`. 
```{r, eval = FALSE}
m3 <- lm(mp ~ disp + cyl + _______, data=new_data) #use * for interaction- include both variables

@@

We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r, eval = FALSE}
#m4:
m4<-lm(_________)

How might we present these results in a table? Both `xtable()` and `stargazer()` are functions that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try both of these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, eval=FALSE, echo=TRUE}
#These produce code for a .pdf that we can copy and paste
xtable(______)
stargazer(_______)

#you can also add lots of useful information onto these
#check out guides online!

When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object.

Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

```{r}
#fitting our model
m5 <- lm(mpg ~ disp + cyl, data = mtcars)

#looking at our model
summary(m5)

@@

#what are we doing here?
colnames(new_dat_n_1) <- c("disp", "cyl")
colnames(new_dat_n_2) <- c("disp", "cyl")
colnames(new_dat_n_3) <- c("disp", "cyl")
#we just created three data frames - all with our x_vec_n vaalues for "disp"
#and three values we have set for "cyl"

@@

#ggplot is rather particular about the format of the data you will be using
plot_dat_n <- data.frame(fit_n_1, fit_n_2, fit_n_3, new_dat_n_1)

#note we never call `cyl` on its own below so the iteration of `new_dat_n_x` we include
#here doesn't matter a ton, it just gives us all values of disp

#this looks like a monster plot, but there are just a lot of layers. A lot of 
#ggplot plot code will end up looking like this
ggplot(data=plot_dat_n) + 
  labs(x = "Displacement of Vehicle", y = "Miles Per Gallon")+
  geom_line(aes(x = disp, y = fit_n_1[, 1], color = "4 cylinders"),linetype = "solid") +
  geom_line(aes(x = disp, y = fit_n_2[, 1], color = "6 cylinders"),linetype = "solid") +
  geom_line(aes(x = disp, y = fit_n_3[, 1], color = "8 cylinders"), linetype = "solid") +

  #what are we doing with the ribbons? What does alpha do? Play around with it.
  geom_ribbon(aes(x = disp, ymin = fit_n_1[, 2], ymax = fit_n_1[, 3]), alpha = .3, fill = "red") +
  geom_ribbon(aes(x = disp, ymin = fit_n_2[, 2], ymax = fit_n_2[, 3]), alpha = .3, fill = "green") +
  geom_ribbon(aes(x = disp, ymin = fit_n_3[, 2], ymax = fit_n_3[, 3]), alpha = .3, fill = "blue") +
  theme(legend.title = element_text(size = 14),
        legend.position = "bottom",
        panel.grid.minor.x = element_blank(), #what does this do?
        panel.grid.minor.y = element_blank()) +
  scale_color_discrete(name="Number of Cylinders") + #what does this seem to do? what happens if you take it out?
  theme_bw()
```

@@

all(fit_n_1 == fit_n_1_a) #what does all() do here?

@@

# Part 3: LM practice
Using the palmerpenguines dataset, create an OLS model using multiple predictors and then display the results in a publication quality table (using stargazer).
```{r}
library(palmerpenguins)
library(stargazer)

#Check out the data file
(penguins)

#Create OLS file

model1<-lm(______,data=penguins)

#Create table

stargazer(____)
```

>>>>

---
title: "Lab 3: The Linear Model in R"
author: "Eric Parajon"
output: pdf_document
---

Adapted  from code and text from Isabel Laterzo, Simon Hoellerbauer and Chelsea Estancona.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = FALSE, echo=TRUE)

@@

```{r}
set.seed(123)

##Explanatory variables
x1 <- rnorm(20, 1.3, 2.3)

@@

error <- rnorm(length(x1), 0, sigma) #Let's simulate some random error, too. 
#Random error is always included in a regression model because unless there is a linear relationship between the variables in the model, there will always be some amount of error (or variation) which is not totally unaccounted for by the model.
```

What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r}
y <- b0 + b1 * x1 + error
```

@@

```{r}
plot(x1, y)
```

Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. The basic syntax for this function is `lm(Y ~ X1 + X2)`. Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

```{r}
m1 <- lm(y ~ x1) #the basic syntax here: lm(Y ~ X1 + X2)
#summarize the 'model object' we've created
summary(m1)
```

@@

```{r}
#examine data
summary(mtcars)
```

@@

Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).

```{r}

new_data <- mtcars %>%
  filter(cyl > 4) %>%
  select(cyl, disp, mpg)
```

Now that we have our data ready, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. The basic syntax for this function is `lm(Y ~ X1 + X2)`. Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

@@

```{r}
#model
m2 <- lm(mpg~disp+cyl, data=new_data)

#summarize
summary(m2)

@@

```{r, eval=FALSE, echo=TRUE}
#Taking a look at m2
View(m2)

#extract coefficients
m2$coefficients

#extract residuals
m2$residuals

#extract y hat 
m2$fitted.values
```

@@

```{r}
m2$coefficients[1] #What does this give us?
m2[2] #What about this?
```

What if we wanted to include an interaction term?
```{r}
m3<-lm(mpg~disp+cyl+disp*cyl, data=new_data) #use * for interaction- include both variables

@@

We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r}
#m3:
m4 <- lm(mpg ~ disp*cyl, data=new_data)

summary(m4)
```

How might we present these results in a table? Both `xtable()` and `stargazer()` are functions that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try both of these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, eval=FALSE, echo=TRUE}

xtable(m4)
stargazer(m4)

#you can also add lots of useful information onto these
#check out guides online!
```

When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object. Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

```{r}
#fitting our model
m5 <- lm(mpg ~ disp + cyl, data = mtcars)
#looking at our model
summary(m5)

@@

#what are we doing here?
colnames(new_dat_n_1) <- c("disp", "cyl")
colnames(new_dat_n_2) <- c("disp", "cyl")
colnames(new_dat_n_3) <- c("disp", "cyl")
#we just created three data frames - all with our x_vec_n values for "disp"
#and three values we have set for "cyl"

@@

#ggplot is rather particular about the format of the data you will be using
plot_dat_n <- data.frame(fit_n_1, fit_n_2, fit_n_3, new_dat_n_1)
#note we never call `cyl` on its own below so the iteration of `new_dat_n_x` we include
#here doesn't matter a ton, it just gives us all values of disp

#this looks like a monster plot, but there are just a lot of layers. A lot of 
#ggplot plot code will end up looking like this
ggplot(data=plot_dat_n) + 
  labs(x = "Displacement of Vehicle", y = "Miles Per Gallon")+
  geom_line(aes(x = disp, y = fit_n_1[, 1], color = "4 cylinders"),linetype = "solid") +
  geom_line(aes(x = disp, y = fit_n_2[, 1], color = "6 cylinders"),linetype = "solid") +
  geom_line(aes(x = disp, y = fit_n_3[, 1], color = "8 cylinders"), linetype = "solid") +
  #what are we doing with the ribbons? What does alpha do? Play around with it.
  geom_ribbon(aes(x = disp, ymin = fit_n_1[, 2], ymax = fit_n_1[, 3]), alpha = .3, fill = "red") +
  geom_ribbon(aes(x = disp, ymin = fit_n_2[, 2], ymax = fit_n_2[, 3]), alpha = .3, fill = "green") +
  geom_ribbon(aes(x = disp, ymin = fit_n_3[, 2], ymax = fit_n_3[, 3]), alpha = .3, fill = "blue") +
  theme(legend.title = element_text(size = 14),
        legend.position = "bottom",
        panel.grid.minor.x = element_blank(), #what does this do?
        panel.grid.minor.y = element_blank()) +
  scale_color_discrete(name="Number of Cylinders") + #what does this seem to do? what happens if you take it out?
  theme_bw()
```

@@

all(fit_n_1 == fit_n_1_a) #what does all() do here? #check in a vector whether all the values are true or not.

@@

# Part 3: LM practice
Using the palmerpenguines dataset, create an OLS using multiple predictors and then display the results in a publication quality table (using stargazer).
```{r}
library(palmerpenguins)
library(stargazer)

#Running model
bill_model <- lm(bill_length_mm ~ species+sex+body_mass_g, penguins)

summary(bill_model)

#Putting results in table
stargazer(bill_model,title="Effect of Predictors on Bill Length",digits=3,star.cutoffs = c(0.1,0.05,0.01),
          label="bill_model",
          covariate.labels = c("\\textbf{Species} (reference= Adélie) &   \\\\
                      \\hspace{1cm}Chinstrap",
                      "\\hspace{1cm}Gentoo",
                    "\\textbf{Sex} (reference= Female) &   \\\\
                      \\hspace{1cm}Male",
                      "\\textbf{Body mass (grams)}"),
          notes="\\parbox[t]{12cm}{Coefficients reported from OLS regression models. Significance codes:*$<$ 0.10,**p$<$ 0.05, ***p$<$ 0.01, two-tailed tests.}",
          notes.append=FALSE,
          style="apsr",
          dep.var.caption = c(""),
          dep.var.labels.include = FALSE,
          table.layout ="-c-!t-s-n",
          header=FALSE,
          notes.align= "l",
          column.sep.width = "1pt",
          font.size = "scriptsize",
          no.space = TRUE)

### 2023-01-04 Lab 4  Graphical Interpretation of LMs/Lab4_Graphical_interpretation_LM_Answers.Rmd
Add files via upload

#Part 1: Ggplot visualizations
Lets begin by returning to the mtcars data set and creating a basic scatterplot of displacement (disp) vs mpg.

```{r, message=FALSE, warning=FALSE}
carsplot <- ggplot(data = mtcars, aes(x = ____, y = ____)) + 
  ____(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()

@@

#Okay great so there's an effect! But what about confounders? Perhaps the penguins species is also important.

@@

@@

#Part 3: LM practice
Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). Then create a model including an additional predictor (of your choice). Create a coefficient plot of the results of the two models and compare them.

#Creating model 1
model1 <- lm(______)

#Tidying model 1
model1_tidy<-______
  mutate(model = "Model 1")

#Creating model 2
model2 <- lm(______)

#Tidying model 2
model2_tidy<-______ %>% 
  mutate(model = "Model 2")

#Combine results into a single dataframe
sleep_results <- bind_rows(model1_tidy, model2_tidy)

#Plotting (make sure to assign color and shape to the model variable in the ggplot call)
sleep_results_plot <- ggplot(_____)

#Interpret 
```

>>>>

#Part 1: Ggplot visualizations
Lets begin by returning to the mtcars data set and creating a basic plot of Displacement vs mpg.

```{r, message=FALSE, warning=FALSE}
carsplot <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()

#One method for saving plots:
#pdf("Plot1.pdf")
#carsplot
#dev.off()

#Where do these plots go?!

#Unique to ggplot:
#ggsave("Plot1a.pdf") #can edit image sizes as well
```

@@

#Okay great so huge effect! But what about confounders? Perhaps the penguins species is also important.

@@

```
#Part 3: LM practice
Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). Then create a model including an additional predictor (of your choice). Create a coefficient plot of the results of the two models and compare them.

#Creating model 1
model1 <- lm(sleep_rem ~ bodywt + brainwt, data = msleep)

#Tidying model 1
model1_tidy<-broom::tidy(model1, se = 'standard', conf.int = TRUE,conf.level = 0.95) %>% 
  mutate(model = "Model 1")

#Creating model 2
model2 <- lm(formula = sleep_rem ~ bodywt + brainwt + sleep_total, data = msleep)

#Tidying model 2
model2_tidy<-broom::tidy(model2, se = 'standard', conf.int = TRUE,conf.level = 0.95) %>% 
  mutate(model = "Model 2")

#Combine results into a single dataframe
sleep_results <- bind_rows(model1_tidy, model2_tidy) %>% 
  filter(term!="(Intercept)")

#Plotting (make sure to assign color and shape to the model variable in the ggplot call)
sleep_results_plot <- ggplot(data=sleep_results, aes(x=term, y=estimate, 
           color = model, shape = model)) + 
    geom_hline(yintercept=0, color="red", size=.5) +
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, width=0), size=.5, position=position_dodge(width=0.5)) +
    geom_point(aes(y=estimate), size=1.75, position = position_dodge(width=0.5)) +
  labs(title = "Model Estimates of Brain and Body Weight (and total sleep) on REM Sleep",
       x = "Predictor",
       y = "Coefficient Estimate",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.",
       color = "Model:",
       shape = "Model:") +
   scale_x_discrete(labels = c("Body Weight", "Brain Weight", "Total Sleep"))+ 
  coord_flip()+
  plot_theme

@@

### 2023-01-04 Lab 5  Optim/Lab5_Optim_AnswerKey.Rmd
Add files via upload

---
title: "Lab 5: Optimization"
author: "Eric Parajon"
output: pdf_document
---

@@

pacman::p_load(tidyverse,broom)
```

Some of today's code has been adapted from Simon Hoellerbauer, Isabel Laterzo and Sean Norton.

@@

Let's suppose we're "hiking" up that mathematical hill trying to find the maximum, and using the first derivative to decide how big of a "step" we take from our current location. However, the first derivative only gives us information on the steepness of that hill. If all we're paying attention to is gaining elevation, we could take a very circular path up to the top. That's where the second derivative comes in.

@@

Here, we are going to set a starting point for the NR method of $\mu_0 = 10$. We will also set our iterations to 7, and then plot the score function with the values for each of these iterations. Don't worry so much about this code, but look at the plot. 
```{r, include = TRUE, message = FALSE, warning = FALSE}

#first generate our data (remember to set the seed)

set.seed(123)
x <- rpois(100, 5)
xbar <- mean(x)
n <- length(x)

@@

Here we can see that across the 7 iterations, we get closer and closer to the root (which is 5.1). The algorithm continues (for as many iterations as you specify) until the parameter values stabilize, or $\theta_{n+1}$ and $\theta_{n}$ become approximately equal, and the root is approximated. If this doesn't make sense to you, watch \href{https://www.youtube.com/watch?v=-5e2cULI3H8}{this video}.

We can see that by the 7th NR iteration, we are quite close to the root (5.1).This method is good for finding local optima as well, if you have a multimodal case for example. In those cases, you can specify intervals and find the local optimum (minimum or maximum).

@@

  logl <- sum(dnorm(e, mean = 0, sigma, log = TRUE)) #using dnorm again

  return(-logl) #we do negative because the optim fn automatically minimizes.
}

@@

```{r, include = TRUE}
set.seed(123)

init_par <- rnorm(3) #values for our parameters, drawn from the normal
                      #just like our data

@@

#did this converge in the min number of iterations?
converge <- p$convergence
converge #yes

@@

In addition, we could do the above process all in one function. Try filling in the function below and optimizing.
````{r, include = TRUE}
#Could do this all in one function
lm.optim <- function(par, y, X, fn = ols.lf){

@@

\newpage
#Practice on your own:
Now, on your own, try doing this with `mtcars` data, instead of generated data. Let's say we want to set our $y$ value as `mpg`, $x1$ as `wt`, and $x2$ as `gear`.

@@

y2 <- mtcars$______
X2 <-cbind(1,______,________)

#in practice, you can use other ways to select your starting values
#such as a grid search, but we will stick with selecting from a normal distribution
set.seed(2030)
params <- rnorm(4)

#now for the optim() part!
p2 <- optim(par = ____, #our starting parameters

           _____, #the function we want it to optimize

           ______, #return Hessian matrix (need for se)

           _______, #calls a quasi-Newton method

@@

#we need hessian = T for the se
coef2 <- ____
_____

#the final element is the log of sigma2 
rse2 <- ____ #standard deviation of conditional distribution
____

se2_2 <- _____ #se
____

vals <- _____ #log likelihood min
____

#did this converge in the min number of iterations?
converge <- ______
_________

#compare to lm() model
m2 <- lm(mpg ~________
__________

>>>>

---
title: "Lab 5: Optimization: Answers"
author: "Eric Parajon"
output: pdf_document
---

@@

#Installing and loading packages for use
pacman::p_load(tidyverse,broom)
```

@@

Let's suppose we're "hiking" up that mathemetical hill trying to find the maximum, and using the first derivative to decide how big of a "step" we take from our current location. However, the first derivative only gives us information on the steepness of that hill. If all we're paying attention to is gaining elevation, we could take a very circular path up to the top. That's where the second derivative comes in.

@@

Here, we are going to set a starting point for the NR method of $\mu_0 = 10$. We will also set our iterations to 7, and then plot the score function with the values for each of these iterations. Don't worry so much about this code, but look at the plot.

```{r, include = TRUE, message = FALSE, warning = FALSE}

#first generate our data (remember to set the seed)
set.seed(123)
x <- rpois(100, 5) #sample from a poisson
xbar <- mean(x) #mean of the poisson samples
n <- length(x) #n obs

@@

Here we can see that across the 7 iterations, we get closer and closer to the root (which is 5.1). The algorithm continues (for as many iterations as you specify) until the parameter values stabilize, or $\theta_{n+1}$ and $\theta_{n}$ become approximately equal, and the root is approximated. If this doesn't make sense to you, watch \href{https://www.youtube.com/watch?v=-5e2cULI3H8}{this video}. We can see that by the 7th NR iteration, we are quite close to the root (5.1).This method is good for finding local optima as well, if you have a multimodal case for example. In those cases, you can specify intervals and find the local optimum (minimum or maximum).

@@

  logl <- sum(dnorm(e, mean = 0, sigma, log = TRUE)) #using dnorm again 
                                               #but note the differences

  return(-logl) #we do negative because the optim fn automatically minimizes.
}
```

@@

```{r, include = TRUE}
set.seed(123118)

init_par <- rnorm(3) #values for our paramters

@@

#did this converge in the min number of iterations?
converge <- p$convergence
converge #yes
```

@@

In addition, we could do the above process all in one function. Try filling in the function below and optimizing.

#Could do this all in one function
lm.optim <- function(par, y, X, fn = ols.lf){

@@

# Practice on your own:
Now, on your own, try doing this with `mtcars` data, instead of generated data. Let's say we want to set our $y$ value as `mpg`, $x1$ as `wt`, and $x2$ as `gear`.

@@

y2 <- mtcars$mpg
X2 <-cbind(1, mtcars$wt, mtcars$gear)

#in practice, you can use other ways to select your starting values
#such as a grid search, but we will stick with selecting from a normal distrib
set.seed(2030)
params <- rnorm(4)

p2 <- optim(par = params, #our starting parameters

@@

#we need hessian = T for the se
coef2 <- p2$par #contains estimated arg mins
coef2

#the final element is the log of sigma2 
rse2 <- exp(coef2[4]) #standard deviation of conditional distribution
rse2

se2_2 <- sqrt(diag(solve(p2$hessian))) #se
se2_2

vals2 <- p2$value #log likelihood min
vals2

#did this converge in the min number of iterations?
converge2 <- p2$convergence
converge2 #yes

#compare to lm() model
m2 <- lm(mpg ~ wt + gear, data = mtcars)
summary(m2)
```

### 2023-01-04 Lab 6  Diagnostics and simulation based inferences/Lab6_Simulation_Uncertainty_Answers.Rmd
Add files via upload

---
title: "Lab 6 - Diagnostics and simulation based inferences "
author: "Eric Parajon"
output: pdf_document
---

@@

We will be examining **Regression Diagnostics**, or ways to examine if our regression meets necessary assumptions, and if anything weird (e.g., outliers, etc.) is going on. Remember the four key assumptions of a linear regression:

1) Linearity of the data - the relationship between x and y is linear

2) Normality of residuals - the residual errors are normally distributed

3) Homogeneity of residual variance - the residuals have CONSTANT variance (homoscedasticity)

4) Independence of residual error terms

The following tests will help us see if these assumptions hold, and will examine a number of possible issues. These include (but are not limited to) non-linearity of the data, heteroscedasticity (non-constant variance in the errors), and presence of influential values (outliers and high-leverage points).

# Simulation
```{r, message = FALSE}
#clear global environ
rm(list=ls())

@@

#Installing and loading packages for use
pacman::p_load(MASS,#required for multivariate normal
               tidyverse,
               AER) #For guns data

Download the "aid_data_95_ee.csv" in the Labs folder on Sakai and then read it into R:
```{r, warning = F, message = F}
#remember to set your working directory!
setwd("______")

#read_csv comes from readr from the tidyverse suite of packages
aid_data_95_ee <- read_csv("aid_data_95_ee.csv")
#also look into fread()
```

Let's take a look at this data very quickly:
```{r, warning = F, message = F, include = F}
#checking dimensions
dim(aid_data_95_ee)
#that's a lot of variables!

@@

Let's look at the distribution of the outcome variable (with some simple, barebones graphs):
```{r, warning = FALSE}
#distribution of civil society strength
ggplot(data = aid_data_95_ee, aes(x = civ_soc_strength)) + 
  geom_density()
#Is it necessarily a problem that our outcome variable is skewed?

@@

Let's fit our model:
```{r, message = F, warning = F, results = "hide"}
#first we have to make sure that EU_accession is a factor, and that the 
#reference category makes sense for us
aid_data_95_ee$EU_accession <- factor(aid_data_95_ee$EU_accession) %>%
  relevel(ref = "no leverage")
levels(aid_data_95_ee$EU_accession) #First one will be the reference catagory

@@

We are now going to use the "average value approach" to simulating predicted outcomes and uncertainty around them. We want to see how predicted civil society strength varies according to lagged civil society assistance. In order to do this, there are a few thing we have to do first:
```{r, results = "hide"}
# we are going to let lagged log civil society assistance vary along its 
# inner-quartile range. These will be the values over which we plot the predicted
# civil society strength
cs_aid_sim <- with(aid_data_95_ee,
                   #quantile() here produces sample quantities based on the given
                   #probability we provide it, here 0.25
                   seq(quantile(civ_soc_aid_l1_log, 0.25, na.rm = T),
                       quantile(civ_soc_aid_l1_log, 0.75, na.rm = T),
                       length.out = 100))

# we now need to create a matrix of hypothetical predictors. We will use the 
# mean of lagged log GDP per capita; in order to see the effect of EU accession
# leverage, we will also allow that to vary (otherwise, you usually use the mode
# of categorical variables)
x_mat_sim <- with(aid_data_95_ee,
                  model.matrix(~ GDPpc_ppp_WB_2011_l1_log + 
                                 EU_accession*civ_soc_aid_l1_log,
                               data = data.frame(GDPpc_ppp_WB_2011_l1_log = 
                                        mean(GDPpc_ppp_WB_2011_l1_log, 
                                             na.rm = T), #mean of lagged log GDP
                                      EU_accession = relevel(
                                        factor(c("no leverage",
                                                  "active", # vary EU accession
                                                  "member")), 
                                        ref = "no leverage"),
                                      #calling above data for civ soc assist
                                      civ_soc_aid_l1_log = rep(cs_aid_sim,
                                                               each = 3))))

@@

#how does it look? note the interaction effects included!
head(x_mat_sim)

@@

dim(cs_str_hat) #300 x 500

@@

#Creating data for rug plot
cs_aid_IQ <- dplyr::select(aid_data_95_ee, civ_soc_aid_l1_log, EU_accession) %>%
  filter(civ_soc_aid_l1_log > quantile(civ_soc_aid_l1_log, 0.25, na.rm = T) &
           civ_soc_aid_l1_log < quantile(civ_soc_aid_l1_log, 0.75, na.rm = T)) %>%
  rbind(data.frame(civ_soc_aid_l1_log = rep(NA, nrow(pred_df) - nrow(.)),
                  EU_accession = rep(NA, nrow(pred_df) - nrow(.))))
```
Now let's graph this!

```{r}
#a library with different colors for ggplot2
library("RColorBrewer")

@@

  theme_bw()
  #theme
```
# Regression Diagnostics

@@

# Practice
To do on your own:

@@

```{r, eval = FALSE}
data(Guns)

#estimate model
guns_model <- lm(_________)

set.seed(123)

#sampling betas
samp_beta_guns <- mvrnorm(____, 
                          coef(____),
                          vcov(____))

#confidence intervals
t(apply(____, 2, quantile, prob = c(.1, .9)))
confint(guns_model, ____)

@@

#hypothetical predictor matrix
 X_sim_guns <- with(Guns,
                   model.matrix(~ ____,
                      data = data.frame(male = mean(male),
                                        income = mean(income),
                                        law = c(____),
                                        density = rep(density_sim, each = 2))
                   ))

@@

#we only need quantiles for our confidence intervals
robbery_hat_quant <- apply(robbery_hat, 1,
                      quantile, c(________))

pred_df_guns <- data.frame(PP = robbery_hat_quant[2, ],
                      UB = robbery_hat_quant[3, ],
                      LB = robbery_hat_quant[1, ],
                      Density = X_sim_guns[, "density"],
                      Law = c("No", "Yes")[X_sim_guns[, "law"] + 1])

ggplot(data =____) +
  geom_ribbon(aes(____fill = Law), alpha = 0.5) +
  geom_line()

>>>>

---
title: "Lab 6 - Diagnostics and simulation based inferences: Answers"
author: "TA: Eric Parajon"
output: pdf_document
---

@@

```{r}
#clear global environ
rm(list=ls())

@@

#Installing and loading packages for use
pacman::p_load(MASS,#required for multivariate normal
               tidyverse,
               AER)
```

Download the "aid_data_95_ee.csv" in the Labs folder on Sakai and then read it into R:
```{r, warning = F, message = F}
#remember to set your working directory!

#read_csv comes from readr from the tidyverse
aid_data_95_ee <- read_csv("aid_data_95_ee.csv")
```

Let's take a look at this data very quickly:
```{r}
#checking dimensions
dim(aid_data_95_ee)
#that's a lot of variables!

@@

Let's look at the distribution of the outcome variable (with some simple, barebones graphs):
```{r}
#distribution of civil society strength
ggplot(data = aid_data_95_ee, aes(x = civ_soc_strength)) + 
  geom_density()
#Is it necessarily a problem that our outcome variable is skewed?

@@

Let's fit our model:
```{r}
#first we have to make sure that EU_accession is a factor, and that the 
#reference category makes sense for us
aid_data_95_ee$EU_accession <- factor(aid_data_95_ee$EU_accession) %>%
  relevel(ref = "no leverage")
levels(aid_data_95_ee$EU_accession) #First one will be the reference catagory

@@

We are now going to use the "average value approach" to simulating predicted outcomes and uncertainty around them. We want to see how predicted civil society strength varies according to lagged civil society assistance. In order to do this, there are a few thing we have to do first:
```{r}
# we are going to let lagged log civil society assistance vary along its 
# inner-quartile range. These will be the values over which we plot the predicted
# civil society strength
cs_aid_sim <- with(aid_data_95_ee,
                   #quantile() here produces sample quantities based on the given
                   #probability we provide it, here 0.25
                   seq(quantile(civ_soc_aid_l1_log, 0.25, na.rm = T),
                       quantile(civ_soc_aid_l1_log, 0.75, na.rm = T),
                       length.out = 100))

# we now need to create a matrix of hypothetical predictors. We will use the 
# mean of lagged log GDP per capita; in order to see the effect of EU accession
# leverage, we will also allow that to vary (otherwise, you usually use the mode
# of categorical variables)
x_mat_sim <- with(aid_data_95_ee,
                  model.matrix(~ GDPpc_ppp_WB_2011_l1_log + 
                                 EU_accession*civ_soc_aid_l1_log,

                               data = data.frame(GDPpc_ppp_WB_2011_l1_log = 
                                        mean(GDPpc_ppp_WB_2011_l1_log, 
                                             na.rm = T), #mean of lagged log GDP

                                      EU_accession = relevel(
                                        factor(c("no leverage",
                                                  "active", # vary EU accession
                                                  "member")), 
                                        ref = "no leverage"),

                                      #calling above data for civ soc assist
                                      civ_soc_aid_l1_log = rep(cs_aid_sim,
                                                               each = 3))))

@@

#how does it look? note the interaction effects included!
head(x_mat_sim)
```

@@

dim(cs_str_hat)

@@

#for rug plot
cs_aid_IQ <- dplyr::select(aid_data_95_ee, civ_soc_aid_l1_log, EU_accession) %>%
  filter(civ_soc_aid_l1_log > quantile(civ_soc_aid_l1_log, 0.25, na.rm = T) &
           civ_soc_aid_l1_log < quantile(civ_soc_aid_l1_log, 0.75, na.rm = T)) %>%
  rbind(data.frame(civ_soc_aid_l1_log = rep(NA, nrow(pred_df) - nrow(.)),
                  EU_accession = rep(NA, nrow(pred_df) - nrow(.))))
```

Now let's graph this!
```{r}
#a library with different colors for ggplot2
library("RColorBrewer")

@@

  theme_bw()
  #theme
```

# Regression Diagnostics

@@

# To do on your own:

@@

```{r}
library(AER)
data(Guns)

#estimate model
guns_model <- lm(robbery ~ male + income + density*law,
                 data = Guns)

set.seed(53)

#sampling betas
samp_beta_guns <- mvrnorm(10000, 
                          coef(guns_model),
                          vcov(guns_model))

#confidence intervals
t(apply(samp_beta_guns, 2, quantile, prob = c(.1, .9)))
confint(guns_model, level = .8)

@@

#hypothetical predictor matrix
 X_sim_guns <- with(Guns,
                   model.matrix(~ male + income + density*law,
                      data = data.frame(male = mean(male),
                                        income = mean(income),
                                        law = c(0, 1),
                                        density = rep(density_sim, each = 2))
                   ))

@@

#we only need quantiles for our confidence intervals
robbery_hat_quant <- apply(robbery_hat, 1, #why 1?
                      quantile, c(0.1, 0.5, 0.9))

pred_df_guns <- data.frame(PP = robbery_hat_quant[2, ],
                      UB = robbery_hat_quant[3, ],
                      LB = robbery_hat_quant[1, ],
                      Density = X_sim_guns[, "density"],
                      Law = c("No", "Yes")[X_sim_guns[, "law"] + 1]) #why can we do this?  `

ggplot(data = pred_df_guns, aes(x = Density, y = PP, lty = Law)) +
  geom_ribbon(aes(ymin = LB, ymax = UB, fill = Law), alpha = 0.5) +
  geom_line() + #what do you think this does?
  xlab("Population Density") +
  ylab("Predicted Number of Robberies") +
  theme_bw()
